### 2024-03-04 presentation/parts/overview_presentation.tex
WIP

\section{Overview}

@@

    \note[item]<1->{So how does the thesis tackle this problem of building such a system?}
    \note[item]<2->{First we examine how a certain class of generative models can be used to detect data that is different from training data.}
    \note[item]<2->{In our second paper we generalize this to be model-agnostic and phrased as a statistical test.}
    \note[item]<3->{We then take a step back and provide an overview of modern methods for unsupervised speech representation learning.}
    \note[item]<4->{We then benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.}
    \note[item]<5->{We then move to a different domain, medical coding, and provide a critical review of recent works.}
    \note[item]<6->{Finally, we provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.}
    \end{frame}

>>>>

\section*{overview}

@@

    \note[item]{So how does the thesis tackle this problem of building such a system?}
    \note{
        \begin{itemize}
            \item [Chp. 4] Examine how a certain class of generative models can be used to detect data that is different from training data. 
            \item [Chp. 5] Generalize this to be model-agnostic and phrased as a statistical test.
            \item [Chp. 6] Step back and provide an overview of modern methods for unsupervised speech representation learning.
            \item [Chp. 7] Benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.
            \item [Chp. 8] Move to a different domain, medical coding, and provide a critical review of recent works.
            \item [Chp. 9] Provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.
        \end{itemize}
    }
    \end{frame}

### 2024-03-04 presentation/parts/presentation_hierarchical.tex
WIP

          % edges
          \edge[]{z3_gen}{z2_gen};
          \edge[]{z2_gen}{z1_gen};
          \edge[]{z1_gen}{x_gen};
          \edge[dashed, bend left]{z2_gen}{x_gen};
          \edge[dashed, bend left]{z3_gen}{x_gen};
          }
          \end{figure}
      \end{column}
  \end{columns}
\end{frame}

@@

\frame{
    \frametitle{What do the lowest latent variables code for?}

@@

% \section{The $\mathcal{L}^{>k}$ likelihood bound}

\frame{
    \frametitle{An alternative likelihood bound, $\mathcal{L}^{>k}$}
    An alternative version of the ELBO that only partially uses the approximate posterior can be written as \cite{maaloe_biva_2019}
    \begin{equation}
        \mathcal{L}^{>k}(\xb; \theta, \phi) = \mathbb{E}_{p_\theta(\zb_{\leq k}|\zb{>k})q_\phi(\zb_{>k}|\xb)} \left[ \log \frac{p_\theta(\xb|\zb)p_\theta(\zb_{>k})}{q_\phi(\zb_{>k}|\xb)} \right]
    \end{equation}

    Here, we have replaced the approximate posterior $q_\phi(\zb|\xb)$ with a different proposal distribution that combines part of the approximate posterior with the conditional prior, namely

    $$p_\theta(\zb_{\leq k}|\zb_{>k})q_\phi(\zb_{>k}|\xb)$$

    This bound uses the conditional prior for the lowest latent variables in the hierarchy.

    \note[item]{So can we come up with a new bound that does not use the lowest latent variables in the same way?}
    \note[item]{So we could use this for OOD detection (as done in BIVA).}
}

% \section{Likelihood ratio}

\frame{
    \frametitle{Likelihood ratios}
    We can use our new bound to compute the score used in a standard likelihood ratio test \cite{buse_likelihood_1982}.
    \begin{equation}\label{eq:llr-as-difference-in-likelihoods}
        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
    \end{equation}
    We can inspect what this likelihood-ratio measures by considering the exact form of our bounds.
    \begin{align}
        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \notag \ .
    \end{align}
    In the likelihood ratio the reconstruction terms cancel out and only the KL-divergences from the approximate to the true posterior remain.
    \begin{align}\label{eq:llr-as-kls}
        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \\
                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \ . \notag
    \end{align}

@@

\begin{frame}
    \frametitle{Conclusions}
    \begin{itemize}
        \item Key observations: 
        \begin{itemize}
            \item The likelihood of a generative model is not a good score for OOD detection \cite{nalisnick_deep_2019}.
            \item Strong correlations between some latent variables for different datasets.
            % \item Reconstructions of OOD data are good when using full approximate posterior.
        \end{itemize}
        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy and showed its effectiveness.
        % \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.
    \end{itemize}
\end{frame}

>>>>

          % edges
          \edge[]{z3_gen}{z2_gen};
          \edge[]{z2_gen}{z1_gen};
          \edge[]{z1_gen}{x_gen};
          \edge[dashed, bend left]{z2_gen}{x_gen};
          \edge[dashed, bend left]{z3_gen}{x_gen};
          }
          \end{figure}
      \end{column}
  \end{columns}

  \note[item]{We want to explain why HVAEs fail at OOD detection when using their likelihood.}
    \note[item]{And how to fix them.}
\end{frame}

@@

\frame{
    \frametitle{What do the lowest latent variables represent?}

@@

\frame{
    \frametitle{Likelihood ratios}
    We suggest to define a likelihood ratio score \cite{buse_likelihood_1982} using the ELBO $\mathcal{L}(\xb)$ and a relaxed bound $\mathcal{L}^{>k}(\xb)$.
    %
    \begin{equation}
        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ ,
    \end{equation}
    %
    where the exact form of the bounds is,
    %
    \begin{align}
        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \notag \ .
    \end{align}
    %
    In the likelihood ratio $log p_\thetab(\xb)$ cancels out and only the KL-divergences from the approximate to the true posterior remain.
    %
    \begin{align}
        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \\
                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \ . \notag
    \end{align}

@@

\begin{frame}
    \frametitle{Conclusions}
    \begin{itemize}
        \item Key observations: 
        \begin{itemize}
            \item The likelihood of a generative model is not a good score for OOD detection \cite{nalisnick_deep_2019}.
            \item Strong correlations between some latent variables for different datasets.
            % \item Reconstructions of OOD data are good when using full approximate posterior.
        \end{itemize}
        \item[\highlight{$\rightarrow$}] Provide explanation for why the likelihood fails fro OODD for HVAEs.
        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the bottom-most latent variables in the hierarchy and showed its effectiveness.
        % \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.
    \end{itemize}
\end{frame}

### 2024-03-04 presentation/parts/presentation_introduction.tex
WIP

\begin{frame}
    \frametitle{What is healthcare?}
    \begin{quotation}
        \noindent\centering
        \textbf{Healthcare} is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
    \end{quotation}
    \begin{flushright}
        \footnotesize
        --- Wikipedia
    \end{flushright}

    \note[item]{Machine learning for healthcare touches on most of these aspects. Fx:}
    \note[item]{
        \begin{itemize}
            \item \textbf{Prevention} by predicting risk of disease allowing for early intervention.
            \item \textbf{Diagnosis} by detecting stroke in emergency calls.
            \item \textbf{Treatment} by suggesting medication.
        \end{itemize}
    }
\end{frame}

>>>>

% \begin{frame}
%     \frametitle{What is healthcare?}
%     \begin{quotation}
%         \noindent\centering
%         \textbf{Healthcare} is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
%     \end{quotation}
%     \begin{flushright}
%         \footnotesize
%         --- Wikipedia
%     \end{flushright}

%     \note[item]{Machine learning for healthcare touches on most of these aspects. Fx:}
%     \note[item]{
%         \begin{itemize}
%             \item \textbf{Prevention} by predicting risk of disease allowing for early intervention.
%             \item \textbf{Diagnosis} by detecting stroke in emergency calls.
%             \item \textbf{Treatment} by suggesting medication.
%         \end{itemize}
%     }
% \end{frame}
